Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-21 submission/manuscript.Rmd
abstract edits CGI

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. We then curated a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males. We found KDM5C represses germline-specific transcription by binding to a subset of germline gene promoters in EpiLCs to facilitate DNA methylation of CpG islands. However, germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding. This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development, due to downstream activation by germline transcription factors. These results define KDM5C's role in germline gene suppression and potentially link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.

- Not sure if I like the last sentence, I think we should focus on what we did do versus what we could do in the future
	- Add something that we deepened the characterization of germline gene classes that use different regulatory mechanisms (CpG islands vs not, meiotic vs late-stage, E2F6/MAX vs no)

>>>>

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice that recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. We then curated a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males. We found KDM5C binds to a subset of germline gene promoters in EpiLCs that contain promoter CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. However, CGI-free germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding. This suggests ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression. These results define KDM5C's role in germline gene suppression and link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.
